Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-06-28 paper.tex
.

A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories. Similar to what \todo{cit Quarck and Jaguar} calls \emph{frozen accidents} in any complex system, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. Temporal and cumulative effects have been considered in various geographical subfields and at various geographical scales, including transportation and urban economics, urban geography, and interregional migrations \citep[for e.g.]{White1977,White1978, AllenSanglier1979,Wilson1981,Pumainetal1989,AllenSanglier1981,WeidlichHaag1988,Portugali2000,Wilson2002,Batty2007,AzizAlaouiBertelle2009}. Less studied is the impact of the spatial setting on models dynamics and potential bifurcations.

@@

In order to test for the influence of spatial initial conditions, we need a systematic method to compare phase diagrams. Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic. A solution is to use systematic quantitative procedures. Several potential methods could be used: for example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters). The use of metrics comparing spatial distributions, such as the Earth Movers Distance, or the comparison of transition matrices of the dynamic associated to the potential described by each distribution.

\comment[JR]{just realized there are a bunch of metrics in physical sciences (hydrology, meteorology, etc.) to compare spatial fields : cf http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0178165 - with spatio-temporal empirical orthogonal functions we can have a 3D parameter space for example ?}

\comment[JR]{See also: http://mck.riks.nl/ to get some ideas.}

To keep it simple and as such methodological considerations are auxiliary to the main purpose of this paper, we propose an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability. More formally, the distance is given by

@@

\comment[JR]{for the plan of the results section, I think the best is to organize by application case.}

@@

>>>>

A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories. Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. Temporal and cumulative effects have been considered in various geographical subfields and at various geographical scales, including transportation and urban economics, urban geography, and interregional migrations \citep[for e.g.]{White1977,White1978, AllenSanglier1979,Wilson1981,Pumainetal1989,AllenSanglier1981,WeidlichHaag1988,Portugali2000,Wilson2002,Batty2007,AzizAlaouiBertelle2009}. Less studied is the impact of the spatial setting on models dynamics and potential bifurcations.

@@

In order to test for the influence of spatial initial conditions, we need a systematic method to compare phase diagrams. Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic. A solution is to use systematic quantitative procedures. Several potential methods could be used: for example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters). The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools. Map comparison methods, popular in environmental sciences, provide numeral tools to compare two dimensional fields~\cite{visser2006map}. To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}. To keep it simple and as such methodological considerations are auxiliary to the main purpose of this paper, we propose an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability. More formally, the distance is given by

@@

@@

More precisely, we explore the basic parameter space of the model, which is the population of agents $P$, the minimal ressource capacity

% summary stats
%    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
%0.009408 0.084490 1.266000 0.930400 1.574000 1.858000

% pca of morphological space
% "","PC1","PC2","PC3","PC4"
%"distance",0.762358566609464,-0.260991693298744,0.200656405132039,0.557162237616392
%"entropy",0.601306167355116,0.181706245959277,0.0958379422351422,-0.772158547261002
%"moran",0.0311129390452153,0.912155429075071,0.30114271129527,0.276256268103684
%"slope",0.237217819823539,0.258531718397015,-0.927289147645628,0.130475642169329

%relativedistance_metaparams.pdf
%relativedistance_morphspace.pdf

%phasediagram_id0_maxSugar110.png
%phasediagram_id27_maxSugar110.png

% phase diagrams
%          spAlpha spDiffsteps spDiffusion spGrowth spPopulation
% id=27 : 0.7913103    2.376837   0.1440293 157.4147 4852.746
% id=0 : 2.562398    3.753032   0.1316788 128.4632 4753.983
% maxSugar = 110
